---
title: "Introduction to the CGDS R library"
author: "Anders Jacobsen"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
  html_document:
    toc: yes
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Introduction to the CGDS R library}
%\VignetteKeywords{cgdsr}
%\VignetteDepends{cgdsr}
%\VignettePackage{cgdsr}
-->

```{r knitrSetup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=TRUE)
```

# Introduction

This package provides a basic set of R functions for querying
the Cancer Genomic Data Server (CGDS) hosted by the Computational
Biology Center (cBio) at the Memorial Sloan-Kettering Cancer Center (MSKCC). This
service is a part of the cBio Cancer Genomics Portal, [http://www.cbioportal.org/](http://www.cbioportal.org/).

In summary, the library can issue the following types of queries:

* getCancerStudies(): What cancer studies are hosted on the server?
For example, TCGA glioblastoma or TCGA ovarian cancer.

* getGeneticProfiles(): What genetic profile types are available for
cancer study X? For example, mRNA expression or copy number alterations.

* getCaseLists(): what case sets are available for cancer study X? For
example, all samples or only samples corresponding to a given cancer subtype.

* getProfileData(): Retrieve slices of genomic data.  For
example, a client can retrieve all mutation data for PTEN and EGFR in
TCGA glioblastoma.

* getClinicalData(): Retrieve clinical data (e.g. patient
survival time and age) for a given cancer study and list of cases.

Each of these functions will be briefly described in the following
sections. The last part of this document includes some concrete examples
of how to access and plot the data.

The purpose of this document is to give the reader a quick overview of
the **cgdsr** package. Please refer to the corresponding R manual
pages for a more detailed explanation of arguments and output for each
function.

# The CGDS R interface

## CGDS(): Create a CGDS connection object
Initially, we will establish a connection to the public CGDS
server hosted by Memorial Sloan-Kettering Cancer Center. The function
for creating a CGDS connection object requires the URL of the CGDS
server service, in this case [http://www.cbioportal.org/](http://www.cbioportal.org/), as an argument.

```{r}
library(cgdsr)
# Create CGDS object
mycgds <- CGDS("http://www.cbioportal.org/")
```

The variable **mycgds** is now a CGDS connection object
pointing at the URL for the public CGDS server. This connection object must
be included as an argument to all subsequent interface
calls. Optionally, we can now perform a set of simple tests of the data
returned from the CGDS connection object using the **test** function:

```{r}
# Test the CGDS endpoint URL using a few simple API tests
test(mycgds) 
```

## getCancerStudies(): Retrieve a set of available cancer studies

Having created a CGDS connection object, we can now retrieve a data
frame with available cancer studies using the **getCancerStudies** function: 

```{r}
# Get list of cancer studies at server
# NOTE: Only show the first few with head()
head(getCancerStudies(mycgds)[, c(1,2)])
```

Here we are only showing the first two columns, the cancer study ID and
short name, of the result data frame. There is also a third column,
a longer description of the cancer study. The cancer study ID must be
used in subsequent interface calls to retrieve case lists and genetic
data profiles (see below).

## getGeneticProfiles(): Retrieve genetic data profiles for a specific cancer study
This function queries the CGDS API and returns the available genetic
profiles, e.g. mutation or copy number profiles, stored about a
specific cancer study. Below we list the current genetic profiles for
the TCGA glioblastoma cancer study:

```{r}
head(getGeneticProfiles(mycgds,'gbm_tcga')[,c(1:2)])
```

Here we are only listing the first two columns, genetic profile ID and
short name, of the resulting data frame. Please refer to the R manual
pages for a more extended specification of the arguments and output.


## getCaseLists(): Retrieve case lists for a specific cancer study
This function queries the CGDS API and returns available case lists
for a specific cancer study. For example, within a particular study, only
some cases may have sequence data, and another subset of cases may
have been sequenced and treated with a specific therapeutic protocol.  Multiple 
case lists may be associated with each cancer study, and this method
enables you to retrieve meta-data regarding all of these case
lists. Below we list the current case lists for the TCGA glioblastoma
cancer study:

```{r}
getCaseLists(mycgds,'gbm_tcga')[,c(1:2)]
```

Here we are only listing the first two columns, case list ID and
short name, of the resulting data frame. Please refer to the R manual
pages for a more extended specification of the arguments and output.

## getProfileData(): Retrieve genomic profile data for genes and genetic profiles
The function queries the CGDS API and returns data based on gene(s),
genetic profile(s), and a case list. The function only allows
specifying a list of genes and a single genetic profile, or oppositely
a single gene and a list of genetic profiles. Importantly, the format of the output
data frame depends on if a single or a list of genes was specified in
the arguments. Below we are retrieving mRNA expression and copy number
alteration genetic profiles for the NF1 gene in all samples of the TCGA glioblastoma
cancer study:

```{r}
getProfileData(mycgds, "NF1", c("gbm_tcga_gistic","gbm_tcga_mrna"), "gbm_tcga_all")[c(1:5),]
```

We are here only showing the first five rows of the data frame. In the next example, we are
retrieving mRNA expression data for the MDM2 and MDM4 genes:

```{r}
getProfileData(mycgds, c("MDM2","MDM4"), "gbm_tcga_mrna", "gbm_tcga_all")[c(1:5),]
```

We are again only showing the first five rows of the data frame.

## getClinicalData(): Retrieve clinical data for a list of cases
The function queries the CGDS API and returns available clinical data (e.g. patient
survival time and age) for a given case list. Results are returned in
a data frame with a row for each case and a column for each clinical
attribute. The available clinical attributes are:

* overall_survival_months: Overall survival, in months.
* overall_survival_status: Overall survival status, usually indicated as "LIVING" or "DECEASED".
* disease_free_survival_months: Disease free survival, in months.
* disease_free_survival_status: Disease free survival status, usually indicated as "DiseaseFree" or "Recurred/Progressed".
* age_at_diagnosis: Age at diagnosis.

Below we retrieve clinical data for the TCGA ovarian cancer dataset (only first five
cases/rows are shown):

```{r}
getClinicalData(mycgds, "ova_all")[c(1:5),]
```

# Examples

## Example 1: Association of NF1 copy number alteration and mRNA expression in glioblastoma
As a simple example, we will generate a plot of the association between
copy number alteration (CNA) status and mRNA expression change for the
NF1 tumor suprpressor gene in glioblastoma. This plot is very similar
to Figure 2b in the TCGA research network paper on glioblastoma
(McLendon et al. 2008). The mRNA expression of NF1 has been 
median adjusted on the gene level (by globally subtracting the median expression
level of NF1 across all samples).

```{r NF1plot1, fig.show=TRUE, echo=TRUE}
df <- getProfileData(mycgds, "NF1", c("gbm_tcga_gistic","gbm_tcga_mrna"), "gbm_tcga_all")
head(df)
boxplot(df[,2] ~ df[,1], main="NF1 : CNA status vs mRNA expression", xlab="CNA status", ylab="mRNA expression", outpch = NA)
stripchart(df[,2] ~ df[,1], vertical=T, add=T, method="jitter",pch=1,col='red')
```

Alternatively, the generic **cgdsr** **plot()**
function can be used to generate a similar plot:

```{r NF1plot2, fig.show=TRUE, echo=TRUE}
plot(mycgds, "gbm_tcga", "NF1", c("gbm_tcga_gistic","gbm_tcga_mrna"), "gbm_tcga_all", skin = 'disc_cont')
```

## Example 2: MDM2 and MDM4 mRNA expression levels in glioblastoma
In this example, we evaluate the relationship of MDM2 and MDM4
expression levels in glioblastoma. mRNA expression levels of MDM2 and MDM4 have been 
median adjusted on the gene level (by globally subtracting the median expression
level of the individual gene across all samples).

```{r MDM2plot1, fig.show=TRUE, echo=TRUE}
df <- getProfileData(mycgds, c("MDM2","MDM4"), "gbm_tcga_mrna", "gbm_tcga_all")
head(df)
plot(df, main="MDM2 and MDM4 mRNA expression", xlab="MDM2 mRNA expression", ylab="MDM4 mRNA expression")
```

Alternatively, the generic **cgdsr** **plot()**
function can be used to generate a similar plot:

```{r MDMplot2, fig.show=TRUE, echo=TRUE}
plot(mycgds, "gbm_tcga", c("MDM2","MDM4"), "gbm_tcga_mrna" ,"gbm_tcga_all")
```

## Example 3: Comparing expression of PTEN in primary and metastatic prostate cancer tumors
In this example we plot the mRNA expression levels of PTEN in primary
and metastatic prostate cancer tumors.

```{r PTENplot, fig.show=TRUE, echo=TRUE}
df.pri <- getProfileData(mycgds, "PTEN", "prad_mskcc_mrna", "prad_mskcc_primary")
head(df.pri)
df.met <- getProfileData(mycgds, "PTEN", "prad_mskcc_mrna", "prad_mskcc_mets")
head(df.met)
boxplot(list(t(df.pri),t(df.met)), main="PTEN expression in primary and metastatic tumors", xlab="Tumor type", ylab="PTEN mRNA expression",names=c('primary','metastatic'), outpch = NA)
stripchart(list(t(df.pri),t(df.met)), vertical=T, add=T, method="jitter",pch=1,col='red')
```
